---
title: "Data Analysis Assignment #2 (75 points total)"
author: "Akerhielm, Nathan"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

### Instructions

R markdown is a plain-text file format for integrating text and R code, and creating transparent, reproducible and interactive reports. An R markdown file (.Rmd) contains metadata, markdown and R code "chunks", and can be "knit" into numerous output types. Answer the test questions by adding R code to the fenced code areas below each item. There are questions that require a written answer that also need to be answered. Enter your comments in the space provided as shown below:

***Answer: (Enter your answer here.)*** 

Once completed, you will "knit" and submit the resulting .html document and the .Rmd file. The .html will present the output of your R code and your written answers, but your R code will not appear.  Your R code will appear in the .Rmd file. The resulting .html document will be graded and a feedback report returned with comments.  Points assigned to each item appear in the template.

**Before proceeding, look to the top of the .Rmd for the (YAML) metadata block, where the *title*, *author* and *output* are given. Please change *author* to include your name, with the format 'lastName, firstName.'**

If you encounter issues with knitting the .html, please send an email via Canvas to your TA.

Each code chunk is delineated by six (6) backticks; three (3) at the start and three (3) at the end. After the opening ticks, arguments are passed to the code chunk and in curly brackets. **Please do not add or remove backticks, or modify the arguments or values inside the curly brackets**. An example code chunk is included here: 

```{r exampleCodeChunk, eval = FALSE, echo = TRUE}
# Comments are included in each code chunk, simply as prompts

#...R code placed here

#...R code placed here

```

R code only needs to be added inside the code chunks for each assignment item. However, there are questions that follow many assignment items. Enter your answers in the space provided. An example showing how to use the template and respond to a question follows.

-----

**Example Problem with Solution:**

Use *rbinom()* to generate two random samples of size 10,000 from the binomial distribution. For the first sample, use p = 0.45 and n = 10. For the second sample, use p = 0.55 and n = 10. Convert the sample frequencies to sample proportions and compute the mean number of successes for each sample. Present these statistics.

```{r Example, eval = TRUE, echo = TRUE}

set.seed(123)
sample.one <- table(rbinom(10000, 10, 0.45)) / 10000
sample.two <- table(rbinom(10000, 10, 0.55)) / 10000

successes <- seq(0, 10)

round(sum(sample.one*successes), digits = 1) # [1] 4.5
round(sum(sample.two*successes), digits = 1) # [1] 5.5
```

**Question: How do the simulated expectations compare to calculated binomial expectations?**

***Answer:  The calculated binomial expectations are 10(0.45) = 4.5 and 10(0.55) = 5.5.  After rounding the simulated results, the same values are obtained.***

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1: (5 points) ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}

par(mfrow=c(1,2))
hist(mydata$RATIO, main="Histogram of Ratio", col="salmon")
qqnorm(mydata$RATIO, main="QQ Plot of Ratio", col="blue")
qqline(mydata$RATIO)

kurtosis(mydata$RATIO, excess = FALSE)
#Notice the difference of 3 mentioned in 1(a)
kurtosis(mydata$RATIO, excess=TRUE)
skewness(mydata$RATIO)


```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}

mydata$L_RATIO <- log10(mydata$RATIO)

par(mfrow=c(1,3))
hist(mydata$L_RATIO, main="Histogram of Log of Ratio", xlab="Ratio", col="salmon")
qqnorm(mydata$L_RATIO,xlab="Log of Ratio", main="QQ Plot of Log of Ratio", col="blue")
qqline(mydata$L_RATIO)
boxplot(mydata$L_RATIO~mydata$CLASS, data=mydata, main="Boxplot of 
        Log of Ratio by Class",
        xlab="Class",
        ylab="Log of Ratio",
        col="magenta")

kurtosis(mydata$L_RATIO, excess = FALSE)
#Notice the difference of 3 mentioned in 1(a)
kurtosis(mydata$L_RATIO, excess = TRUE)
skewness(mydata$L_RATIO)

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
non_log<-bartlett.test(mydata$RATIO~mydata$CLASS)
l<-bartlett.test(mydata$L_RATIO~mydata$CLASS)

print(list('ratio:', non_log, 'log ratio:', l))

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: The RATIO histogram is skewed to the right and the QQ plot deviates from the QQ line, indicating that normality is violated. For the L_RATIO plots, however, there histogram is not skewed, and there are only a few minor deviations in the QQ plot from the QQ line, Additionally, L_RATIO displays a homogeneity of variance across class. Therefore,  L_RATIO exhibits better conformance to normality.***


***#### Section 2 (10 points) ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}


anova_1<-aov(L_RATIO~CLASS*SEX, data=mydata)
anova_2<-aov(L_RATIO~CLASS+SEX, data=mydata)
summary(anova_1)
summary(anova_2)
```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer:The interaction term was insignificant, for the F-values for both CLASS and SEX are nearly identical in the two analysis, regardless of whether or not CLASS:SEX was in the model. Therefore, CLASS and SEX have a statistically significant impact on L_RARTIO while the interaction between CLASS and SEX is not statistically significant when it comes to an effect on L_RATIO.***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}

TukeyHSD(anova_2)

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: We see a negative correlation between the coefficients and age classes, i.e. as the age class increases, the coefficients get smaller. This indicates a negative correlation between L_RATIO and CLASS. Furthermore, there is statistical significance across all classes except A2-A1, which has a p-value of 0.69. The results suggest male and female abalones can be combined into a single category labeled as "adults because M-F has a p-value of 0.94 and hence is statistically insignificant***


***####  Section 3: (10 points) ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)


```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}
par(mfrow=c(1,2))
hist(mydata$VOLUME[mydata$TYPE=="ADULT"], main="Histogram of Adult Volumes",
     col="salmon", xlab="Adult Volume")
hist(mydata$VOLUME[mydata$TYPE=="I"], main="Histogram of Infant Volumes",
     col="lightgreen", xlab="Infant Volume")

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: On the left-hand histogram, we see that the plot of the Adult Volumes follows a more normal distribution, and the majority of volume values range from 200 to 600 The histogram on the right of infant volumes is heavily skewed to the right, and the majority of volumes values range from 0 to 250. Thus, it will be difficult to separate infants from adults based only on volume.***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}
mydata$L_VOLUME <- log10(mydata$VOLUME)
mydata$L_SHUCK <- log10(mydata$SHUCK)


a1<-ggplot(data=mydata, aes(x=VOLUME, y=SHUCK))
a2<-ggplot(data=mydata, aes(x=L_VOLUME, y=L_SHUCK))+
  xlab( "Log of Volume")+
           ylab("Log of Shuck")



grid.arrange(
  a1+geom_point(aes(color=CLASS)),
  a2+geom_point(aes(color=CLASS)),
  a1+geom_point(aes(color=TYPE)),
  a2+geom_point(aes(color=TYPE))

)

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: In all four plots, there is a positive correlation between either VOLUME and SHUCK, or Log of VOLUME and Log of SHUCK. A main difference in the logarthmic plots is that there is less variability amongst the points, and we see a stronger relationship linear between Log of VOLUME and Log of SHUCK than in the non-logarthmic plots. Furthermore, there is clearer distinction in the logarthmic plots between the classes and Adults/Infants. Looking at the two class plots, we see a clear progression from red (A1) to purple (A5) in the log plot, but the colors overlap in the non-log plot. The same pattern is true for the Adult/Infant plots in the second row, as the log plots have a clear distinction in the blue (Infant) and red (Adult), while the non-log plot overlaps these two. ***


***####   Section 4: (5 points) ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

model<-lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data=mydata)
summary(model)
```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer:The results indicate that as CLASS increases, there is a stronger decrease in L_SHUCK. This finding aligns with earlier displays in that we can deduce that L_SHUCK increases more significantly for the lower class ages.***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: TYPE is a less important predictor in this regression than CLASS, and a significantly less important predictor than L_VOLUME. These two observations indicate that overall, TYPE may not be a solid, reliable predictor of L_SHUCK. Rather, we should rely more heavily on CLASS than TYPE. ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: (5 points) ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}
par(mfrow=c(1,2))
hist(model$residuals, main="Histogram of Residuals", xlab = "Residuals", 
     col="royalblue")
qqnorm(model$residuals, main="QQ Plot of Residuals", xlab = "Residuals", 
     col="red")
qqline(model$residuals)

kurtosis(model$residuals, excess=FALSE)
kurtosis(model$residuals, excess=TRUE)

skewness(model$residuals)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}

grid.arrange(
  ggplot(data=mydata, aes(x=L_VOLUME, y=model$residuals)) + 
    geom_point(aes(color = CLASS)),
  ggplot(data=mydata, aes(x=L_VOLUME, y=model$residuals)) + 
    geom_point(aes(color = TYPE)),
  ggplot(data=mydata, aes(x=CLASS,y= model$residuals)) +
    geom_boxplot(aes(fill = CLASS)),
  ggplot(data=mydata, aes(x=TYPE, model$residuals)) + 
    geom_boxplot(aes(fill = TYPE))
)


bartlett.test(model$residuals~CLASS, data=mydata)
```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: The plots reveal that the residuals are fairly randomly spread and near zero on both sides. In the scatter plots, we do notice the residuals tend to clump up near the right side of the plots, so there could be some slight improvements to the model, but overall the points are still evenly distributed which is what we want to see from residual plots. In the Bartlett test, the high p-value indicates that there is homogeneity of variance across the residuals. Therefore, L_VOLUME and VOLUME might be useful for harvesting decisions.***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: (5 points) ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
(max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383
##(max_adl_vol <- max(mydata$VOLUME[mydata$TYPE == "A"])) 

# Proportion of adults harvested
h1_a<-sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) /
  total.adults  # [1] 0.2476573

# Add code to calculate the proportion of infants harvested
h1_i<-sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /
  total.infants



# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:

median_infant<-median(mydata$VOLUME[mydata$TYPE == "I"])


# Add code to calculate the proportion of infants harvested
h2_i<-sum(mydata$VOLUME[mydata$TYPE == "I"] > median_infant) /
  total.infants
# Add code to calculate the proportion of adults harvested
h2_a<-sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > median_infant) /
  total.adults
# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.


# Add code to determine the median adult volume:
median_adult<-median(mydata$VOLUME[mydata$TYPE == "ADULT"])


# Add code to calculate the proportion of infants harvested
h3_i<-sum(mydata$VOLUME[mydata$TYPE == "I"] > median_adult) /
  total.infants

# Add code to calculate the proportion of adults harvested
h3_a<-sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > median_adult) /
  total.adults

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}
plot(volume.value,prop.adults, xlab="", ylab="",type="l", xlim=c(0,1000), ylim=c(0,1), col="red")
par(new=T)
plot(volume.value,prop.infants, xlab="", ylab="", type="l", axes=F, col="blue")
title(main="Proportion of Adults and Infants Protected", xlab="Volume", ylab="Proportion")
legend("bottomright", legend=c("Adult","Infant"), lty=1, col=c("red","blue"))
abline(v=median_infant)
text( x=median_infant+15, y=0.3, labels = ("Median Infant=133.8"), srt=-90)
abline(v=median_adult)
text( x=median_adult+15, y=0.3, labels = ("Median Adult=384.6"), srt=-90)
abline(v=max_inf_vol)
text( x=max_inf_vol+15, y=0.3, labels = ("Protect All Infants=526.6"),srt=-90)

```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: These values suggest that the median infant cutoff is much lower than the median adult cutoff. Thus, the chances of harvesting an infant when attempting to harvest an adult is low, which is good because this means the infants are more protected. We want the infants to be protected in their most vulnerable stages of life, and so when harvesting occurs, we want to kill the adults instead of these infants. ***

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

Note on ROC:

There are multiple packages that have been developed to create ROC curves. However, these packages - and the functions they define - expect to see predicted and observed classification vectors. Then, from those predictions, those functions calculate the true positive rates (TPR) and false positive rates (FPR) and other classification performance metrics. Worthwhile and you will certainly encounter them if you work in R on classification problems.
However, in this case, we already have vectors with the TPRs and FPRs. Our adult harvest proportion vector, (1 - prop.adults), is our TPR. This is the proportion, at each possible 'rule,' at each hypothetical harvest threshold (i.e. element of volume.value), of individuals we will correctly identify as adults and harvest. Our FPR is the infant harvest proportion vector, (1 - prop.infants). We can think of TPR as the Confidence level (ie 1 - Probability of Type I error and FPR as the Probability of Type II error. At each possible harvest threshold, what is the proportion of infants we will mistakenly harvest?
Our ROC curve, then, is created by plotting (1 - prop.adults) as a function of (1 - prop.infants). In short, how much more 'right' we can be (moving upward on the y-axis), if we're willing to be increasingly wrong; i.e. harvest some proportion of infants (moving right on the x-axis)?


-----

***#### Section 7: (10 points)  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}
difference<-((1 - prop.adults) - (1 - prop.infants))


plot(volume.value,difference, xlab="Volume", ylab="Proportion", type="l", col="blue",
     main="Difference in Harvest Proportions of Adults and Infants")

abline(v=median_infant)
text( x=median_infant+15, y=0.3, labels = ("Median Infant=133.8"), srt=-90)
abline(v=median_adult)
text( x=median_adult+15, y=0.3, labels = ("Median Adult=384.6"),srt=-90)

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)



plot(volume.value,difference, xlab="", ylab="", type="l", col="blue")

lines(volume.value,smooth.difference, xlab="", ylab="", type="l", col="red")

title(main="Difference in Harvest Proportions of Adults and Infants", xlab="Volume", ylab="Proportion")
legend("topright", legend=c("Non-Smooth","Smooth"), lty=1, col=c("blue","red"))
abline(v=median_infant)
text( x=median_infant+15, y=0.3, labels = ("Median Infant=133.8"), srt=-90)
abline(v=median_adult)
text( x=median_adult+15, y=0.3, labels = ("Median Adult=384.6"),srt=-90)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}
max<-volume.value[which.max(smooth.difference)]

par(mfrow=c(1,2))
plot(volume.value,difference, xlab="", ylab="", type="l", col="blue")
lines(volume.value,smooth.difference, xlab="", ylab="", type="l", col="red")

title(main="Difference in Harvest Proportions of Adults and Infants", xlab="Volume", ylab="Proportion")
abline(v=max)
text( x=max+15, y=0.28, labels = ("Volume=262.1"), srt=-90)


plot(volume.value,prop.adults, xlab="", ylab="",type="l", xlim=c(0,1000), ylim=c(0,1), col="red")
par(new=T)
plot(volume.value,prop.infants, xlab="", ylab="", type="l", axes=F, col="blue")
title(main="Proportion of Adults and Infants Protected", xlab="Volume", ylab="Proportion")
legend("bottomright", legend=c("Adult","Infant"), lty=1, col=c("red","blue"))
abline(v=median_infant)
text( x=median_infant+10, y=0.3, labels = ("Median Infant"), srt=-90)
abline(v=median_adult)
text( x=median_adult+10, y=0.3, labels = ("Median Adult"), srt=-90)
abline(v=max_inf_vol)
text( x=max_inf_vol+10, y=0.3, labels = ("Protect All Infants"), srt=-90)
abline(v=max)
text( x=max+10, y=0.58, labels = ("Max Difference"), srt=-90)

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the vlume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

h4_a<-(1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
h4_i<-(1-prop.infants)[which.max(smooth.difference)]
# OR,
sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332


```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8: (10 points)  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

cut<-volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786



adult_cut<-sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > cut) /
  total.adults
print(adult_cut)

# Add code to calculate the proportion of infants harvested
infant_cut<-sum(mydata$VOLUME[mydata$TYPE == "I"] > cut) /
  total.infants
print(infant_cut)

```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}
plot(volume.value,prop.adults, xlab="", ylab="",type="l", xlim=c(0,1000), ylim=c(0,1), col="red")
par(new=T)
plot(volume.value,prop.infants, xlab="", ylab="", type="l", axes=F, col="blue")
title(main="Proportion of Adults and Infants Protected", xlab="Volume", ylab="Proportion")
legend("bottomright", legend=c("Adult","Infant"), lty=1, col=c("red","blue"))
abline(v=median_infant)
text( x=median_infant+10, y=0.3, labels = ("Median Infant"), srt=-90)
abline(v=median_adult)
text( x=median_adult+10, y=0.3, labels = ("Median Adult"), srt=-90)
abline(v=max_inf_vol)
text( x=max_inf_vol+10, y=0.3, labels = ("Protect All Infants"), srt=-90)
abline(v=max)
text( x=max+10, y=0.58, labels = ("Max Difference"), srt=-90)
abline(v=cut)
text( x=cut+10, y=0.58, labels = ("Zero A1 Infants"), srt=-90)

```


***#### Section 9: (5 points) ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}

pa<-(1-prop.adults)
pi<-(1-prop.infants)

plot(pi, pa, type="l", col="blue", xlab="Infant Harvest Proportion",
     ylab="Adult Harvest Proportion", main="Adult vs. Infant Harvest Proportion")
abline(a=0,b=1)
points(h1_i,h1_a)
text(h1_i+.1,h1_a, "Protect All Infants", cex=0.7)
points(h2_i, h2_a)
text(h2_i+.1,h2_a, "Median Infant Volume", cex=0.7)
points(h3_i, h3_a)
text(h3_i+.1,h3_a, "Median Adult Volume", cex=0.7)
points(infant_cut, adult_cut)
text(infant_cut+.15, adult_cut, "Zero A1 Infants", cex=0.7)
points(h4_i,h4_a)
text(h4_i+0.1, h4_a, "Max Difference", cex=0.7)
```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
auc(pi,pa)


```


***#### Section 10: (10 points) ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	
protect_all<-(h1_i*total.infants+h1_a*total.adults)/(total.infants+total.adults)
m_inf<-(h2_i*total.infants+h2_a*total.adults)/(total.infants+total.adults)
m_a<-(h3_i*total.infants+h3_a*total.adults)/(total.infants+total.adults)
max_diff<-(h4_i*total.infants+h4_a*total.adults)/(total.infants+total.adults)
zero<-(infant_cut*total.infants+adult_cut*total.adults)/(total.infants+total.adults)


h1<-c(max_inf_vol,h1_a,h1_i, protect_all)
h2<-c(median_infant,h2_a,h2_i, m_inf)
h3<-c(median_adult,h3_a,h3_i, m_a)
h4<-c(max, h4_a, h4_i, max_diff)
h5<-c(cut,adult_cut, infant_cut, zero)

harvest<-rbind(h1,h2,h3,h4,h5)
colnames(harvest)<-c("Volume", "True Positive", "False Positive", "Total Harvest")
rownames(harvest)<-c("Protect All Infants", "Median Infants", "Median Adults",
"Max Difference", "Zero A1 Infants")
kable(harvest)

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: The five cutoffs we examined in this assignment range from aggressive to conservative in the following order: median infant, zero A1 infants, max difference, median adults, and protect all infants. Aggressive means the lowest overall preservation rate of abalones which is represented by median infant with a False Positive rate of 0.49 On the other end of the spectrum, most conservative means the highest overall preservation rate of abalones. This is the case with Protect All Infants, with a false positive rate of 0.0. The overall preservation rate is not just the most or fewest number of abalones preserved, however. For example, the max difference cutoff examines the highest adult-to-infant proportion. Thus, a very large number of adults could be harvest, but few infants, in order to maximize this ratio. From a sustainability standpoint, we would want to choose the median adults or protect all infants cutoff, but at the end of the day, there could be business decisions involved, and hence a higher harvest proportion and hence more aggressive cutoff could be required.***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: Based on my analysis in the above essay question, there is not a one-size-fits-all choice for a cutoff, but rather it depends on what the individual(s) are looking for. If sustainability is the top priority, then I would recommend the median adults or protect all infants cutoff. If maximizing profits is at the top of the list, then I would recommend median infants or zero A1 infants. One qualification I would present regarding this analysis is that this is an observational study, and hence we do not have any control over the variables or outcomes. Therefore, we need to be extremely cautious of coming to any sweeping conclusions based on this analysis. If it's necessary to pick a cutoff, I would recommend the zero A1 infants. This has the second highest harvest proportion at 0.67 and at the same time, it protects the youngest class of infants. Thus, this cutoff can appease both the sustainability and harvest sides of the table. Finally, for future abalone studies of this type, I would recommend taking a broader sample. This would help ensure we have a truly representative abalone sample and hence this would help reduce the false positive rates across the board.***