---
title: 'Data Analysis Assignment #1 (50 points total)'
author: Akerhielm, Nathan
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$RINGS,mydata$CLASS)
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer:The nominal variable is SEX and the ordinal variable is CLASS. The rest of the variables (LENGTH, DIAM, HEIGHT, WHOLE, SHUCK, RINGS) are ratio data. Based on the table, we see quite a few outliers for Rings because there are multiple values beyond the Q3+ 1.5*IQR range. This indicates there will be skewness to the right. We also see that for VOLUME, the max value of 995.7 falls outside of the Q3+1.5*IQR range, so VOLUME has an outlier and is skewed to the right.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}

mytable<-table(mydata$SEX, mydata$CLASS)
addmargins(mytable)
barplot(mytable,
        main="Barplot of Abalone Sex and Class",
        legend = c("Female", "Infant", "Male"),
        col=c("magenta", "skyblue", "green"),
        ylab="Number of Records")
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: Based on the barplot, we can see that the sex distribution of abalones is normally distributed and it's also slightly skewed to the right. This makes sense that most life occurs in the early to middle years and declines in the later years. There are more infants in A1 and A2, then a sharp  decline in A4 and A5. This trend in infants makes sense, for infants are typically born in the early stages. There are more males that females in all classes (A1-A5). The gap between males and females is noticeable in A1-A3 but there is only a slight gap between males and females in A4 and A5. This could suggest that males have a shorter lifespan than females.  ***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
work<-mydata[sample(nrow(mydata), size=200),]
plot(work[,2:6])

```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}


ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
  geom_point(aes(color = CLASS)) +
  ggtitle("Abalone Whole Weight vs. Volume") +
  ylab("Whole Weight") +
  xlab("Volume") 


```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}

ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + 
  geom_point(aes(color = CLASS)) +
  ggtitle("Abalone Whole Weight vs. Shuck") +
  ylab("Shuck") +
  xlab("Whole Weight")+
  geom_abline(intercept =0, slope=max(mydata$SHUCK/mydata$WHOLE))
```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: In both plots, we see a positive correlation between WHOLE and either VOLUME in part (a) or SHUCK in part (b). However, there is  more variability amongst the five classes in the plot in part (a). In this plot, the colors from the five different classes are spread throughout the plot, whereas our plot from part (b) shows more distinct groupings of the classes. We see that the older the abalone, the lower the shuck-weight ratio is. This indicates that as abalone get older, they get smaller. This makes intuitive sense, as many organisms and humans decreaase in size as they age. Personally, I've seen this trend in my grandparents who are in their late 80's, and they are shorter now than they were when I was a teenager. ***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
par(mfrow = c(3,3))
hist(mydata$RATIO[mydata$SEX=="F"], main="Female Ratio", xlab="Ratio", col="red")
hist(mydata$RATIO[mydata$SEX=="I"], main="Infant Ratio", xlab="Ratio", col="green")
hist(mydata$RATIO[mydata$SEX=="M"], main="Male Ratio",xlab="Ratio",col="blue")

boxplot(mydata$RATIO[mydata$SEX=="F"], main="Female Ratio", horizontal = TRUE, col="red")
boxplot(mydata$RATIO[mydata$SEX=="I"], main="Infant Ratio", horizontal = TRUE, col="green")
boxplot(mydata$RATIO[mydata$SEX=="M"], main="Male Ratio", horizontal = TRUE, col="blue")

qqnorm(mydata$RATIO[mydata$SEX=="F"], main="Female Ratio", col="red")
qqline(mydata$RATIO[mydata$SEX=="F"])

qqnorm(mydata$RATIO[mydata$SEX=="I"], main="Infant Ratio", col="green")
qqline(mydata$RATIO[mydata$SEX=="I"])

qqnorm(mydata$RATIO[mydata$SEX=="M"], main="Male Ratio", col="blue")
qqline(mydata$RATIO[mydata$SEX=="M"])


```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: To determine if normality has been upheld, we can examine the QQ-plots, and specifically whether the values deviate from the qq-line. In all three plots, there are deviations along the bottom left and top right of the qq-line, indicating normality is likely violated. This makes sense because in the boxplots, we see outliers for Female, Male, and Infant, and there is skewness to the right in all three histograms. Thus, due to these outliers, our suspicion of non-normality is confirmed with the qq plots.   *** 

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}

f_out<-boxplot.stats(mydata$RATIO[mydata$SEX=="F"])$out
i_out<-boxplot.stats(mydata$RATIO[mydata$SEX=="I"])$out
m_out<-boxplot.stats(mydata$RATIO[mydata$SEX=="M"])$out
ratio_out<-c(f_out, i_out, m_out)
kable(mydata[mydata$RATIO %in% ratio_out,])

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: One of my main observations is that "Infant" has the most outlier RATIO values with 8. Since there is still a lot of growth and development going on during the infant stage, it makes sense they have the most outliers. It's also interesting to note that all but 1 outlier comes from Classes A1-A3. Thus, the outliers from Infants and younger classes may be skewing the data as a whole. As a result of this skewness, it may prove challenging to draw conclusions about the dataset as a whole later on in the data analysis process. ***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}

par(mfrow = c(2,2))
boxplot(mydata$VOLUME~mydata$CLASS, data=mydata, main="Boxplot of Volume by Class",
        xlab="Class",
        ylab="Volume")
boxplot(mydata$WHOLE~mydata$CLASS, data=mydata, main="Boxplot of Whole Weight by Class",
        xlab="Class",
        ylab="Whole Weight")

plot(mydata$RINGS,mydata$VOLUME, main="Rings vs. Volume",
        xlab="Rings",
        ylab="Volume")
plot(mydata$RINGS, mydata$WHOLE, main="Rings vs. Whole Weight",
     xlab="Rings",
        ylab="Whole Weight")


```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: Based on these plots, I think Rings, CLass, Volume, and Whole Weight would be fine predictors of ballpark age. We see there is a positive correlation between Class and Volume, as well as Class and Whole Weight. That is, as Class increases, so too do Volume and Weight.This would help give a rough  estimate of abalone age, for we could deduce that older abalone have larger Volumes and Weights, but it would be difficult to determine a specific age or range of ages. Furthermore, there is a positive correlation between number of Rings and Volume/Weight, but once Rings reaches about 15-20, the Weight and Volume level off. Once again, we could gain c a rough idea of the abalone age,  but it would hard to narrow down on an exact number with these variables alone. Especially using Rings greater than 15-20 could lead to misclassification of age as well. ***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}

a<-aggregate(mydata[c("VOLUME", "SHUCK", "RATIO")], FUN=mean, by=mydata[c("SEX", "CLASS")])
v_matrix<-matrix(a$VOLUME, nrow = 3, dimnames = list(list("F", "I", "M"), 
                                                     list("A1", "A2", "A3","A4","A5")))
s_matrix<-matrix(a$SHUCK, nrow = 3, dimnames = list(list("F", "I", "M"), 
                                                     list("A1", "A2", "A3","A4","A5")))

r_matrix<-matrix(a$RATIO, nrow = 3, dimnames = list(list("F", "I", "M"), 
                                                     list("A1", "A2", "A3","A4","A5")))

kable(v_matrix, caption = "Volume Mean by Sex and Class")
kable(s_matrix, caption = "Shuck Mean by Sex and Class")
kable(r_matrix, caption = "Ratio Mean by Sex and Class")
```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
par(mfrow=c(1,3))

with(mydata, interaction.plot(x.factor=CLASS,
                              trace.factor = SEX,
                              response = RATIO,
                              ylab="Ratio",
                              xlab="Class",
                              col = c("blue", "black", "red"),
                            main = "Mean Ratio per Class"
)
)
     
with(mydata, interaction.plot(x.factor=CLASS,
                              trace.factor = SEX,
                              response = VOLUME,
                              ylab="Volume",
                              xlab="Class",
                              col = c("skyblue", "limegreen", "grey"),
                            main = "Mean Volume per Class"
)
)

with(mydata, interaction.plot(x.factor=CLASS,
                              trace.factor = SEX,
                              response = SHUCK,
                              ylab="Shuck",
                              xlab="Class",
                              col = c("darkgreen", "magenta", "orange"),
                            main = "Mean Shuck per Class"
                            
)
)
```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: One question these plot raises is why do females have a higher volume and shuck than males? Since infants are still not fully developed, I understand why females are higher than infants, but it is unclear why they are higher than males in these two categories. Additionally, I am curious as to why have a larger shuck and volume, but their growth rates over time are smaller than infants and mlaes? Is this because females reach their peak development much sooner? Furthermore, why does shuck to volume ratio decrease at the same rate for males and females as age increases? I would expect there to be some difference in these rates, but the slope of the two lines is nearly identical. Finally, we see that Infants have a smaller Volume and Shuck than males and females, but the question is raised of why do they also have a smaller shuck to volume ratio than male and females? ***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
s<-subset(mydata, subset=mydata$RINGS<16)
par(mfrow = c(2, 2))
boxplot(s$VOLUME[mydata$SEX=="I"]~s$RINGS[mydata$SEX=="I"], 
        xlab="Rings",
        ylab="Volume",
        ylim = c(0, 1100),
        main="Volume by Rings for Infants")

boxplot(s$VOLUME[mydata$SEX!="I"]~s$RINGS[mydata$SEX!="I"], 
        xlab="Rings",
        ylab="Volume",
        ylim = c(0, 1100),
        main="Volume by Rings for Adults")

boxplot(s$WHOLE[mydata$SEX=="I"]~s$RINGS[mydata$SEX=="I"], 
        xlab="Rings",
        ylab="Whole Weight",
        ylim = c(0, 400),
        main="Whole Weight by Rings for Infants")

boxplot(s$WHOLE[mydata$SEX!="I"]~s$RINGS[mydata$SEX!="I"], 
        xlab="Rings",
        ylab="Whole Weight",
        ylim = c(0, 400),
        main="Whole Weight by Rings for Adults")
    
```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: In the Whole Weight plots, when comparing infants and adults, we see in both a positive relationship between Rings and Whole Weight There is much more of a spread of this increase in adults than in infants, however. Intuitively, this makes sense because there is only so much infants will weigh, but in adults there is a greater spread among weights. Furthermore, we observe that in both infants and adults, the largest Whole Weight values occur between approximately ring values 9-13, and then there is a slight decrease in Whole Weight after Rings=13. This could imply when the shucking occurs. In the Volume plots, there is a similar trend. For infants and adults, the Volume and Weight plots are skewed to the left, and we see the biggest values for Volues in both Volume and Weight between rings 9 and 13.   ***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: One plausible reason the original study failed is that variables such as Volume, Shuck, and Whole Weight are decent indicators of a ballpark abalone age, but they do not do well at predicting an exact age or range of ages. In question 3(a), we saw outliers for males, females, and infants, and a violation of normality for all 3 in their QQ-plots. As a result, this makes it difficult to use these aforementioned variables to draw conclusions about specific age. The researchers incorrectly classified abalones with 15-20 rings as infants, but this is the same number that some older adult abalone have. Thus, this mistake likely explains why the non-normality and hence why it's to predict age from physical measurements for younger abalones. ***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: One question I would ask is the sample size. If the sample size was fairly large compared to the overall size of the population, I would feel more comfortable accepting them as representative of the sampled population. I would next ask about the assumptions the researchers made. For example, if I saw in the summary statistics different classifications, I would want to know the assumptions behind how the classifications were assigned. A thorough explanation of these assumptions, and also why they were made, would also make me more likely to accept them as representative. Finally, I would want to know what type of sampling method was used. I would hope to see a simple random sample over a less sound method such as convenience sampling. ***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: When analyzing data derived from observational studies, it's important to remember that there is no interference of the research subjects, and so the lack of control over treatment groups makes drawing inferences difficult. You can determine if some of the variables are correlated, but to jump to the conclusion that X causes Y based on an observational study is very likely to be erroneous. Therefore, observational studies are a great exploratory analysis that can be used to determine correlations between variables. ***
